NumPy---Array Programming in Python

### 2020-01-13 body.tex
Edit Proliferation and Discussion

- massive explosion in data science, machine learning, and artificial intelligence

\fixme{NumPy is now omnipresent.  Many packages depend on it.}

- new hardware platforms; large scales of data; computational barriers; ...

\fixme{arrays are a nice idea in principle, but not all problems are that easily
solved; it breaks down when you have very large data sets, when your data
doesn't live on a single machine, when you have to use specialized hardware, or
when several resources need to be combined to solve a single array computation}

This has lead to a proliferation of arrays (Fig.~\ref{fig:array-proliferation}).

- distributed arrays

- GPUs

- computation graph

- sparse arrays

- other languages w/ arrays

@@

NumPy's API has become ubiquitous in scientific computation, but NumPy itself
can never hope to satisfy every single specialized need of the community.
The advent of new technologies, such as GPUs, create opportunities that should be
exploited, but are out of scope for NumPy development.

Still, the widespread adoption of the NumPy API is valuable, as it lowers the
barrier to entry and provides the wider community with a stable array programming interface.

@@

NumPy was initially developed by students, faculty, and researchers in their
spare time to provide a modern array programming library for Python.
They were influenced by their experiences with powerful interactive programming
languages for scientific computing like Yorick \cite{munro1995using} as well
as commercial languages like IDL and Matlab.
Often they developed code to solve their own or their colleagues' problems.

@@

% teaching (neuroimaging paper / etc.?)

>>>>

Recent years have seen a massive explosion in data science, machine learning, and artificial intelligence.  NumPy and its API, underlying almost every tool in the computational toolchain, has become ubiquitous.  Still, NumPy itself could never hope to satisfy every single specialized need of the community; it is limited in its ability to work with very large datasets, datasets split across multiple systems, and computations that require specialized hardware.

Furthermore, pursuing solutions to these challenges are deemed out of scope for NumPy development. The community's efforts to fill the resultant gaps therefore led to a proliferation of arrays (Fig.~\ref{fig:array-proliferation}): distributed arrays, GPU arrays, delayed execution arrays, sparse arrays, foreign language arrays, and more.

@@

Given a rapid proliferation of array implementations, a widespread adoption of the NumPy API remains valuable: it lowers the barrier to entry for newcomers, and provides the wider community with a stable array programming interface.

@@

NumPy was initially developed by students, faculty, and researchers in their spare time to provide a modern array programming library for Python.  They were influenced by their experiences with powerful interactive programming languages for scientific computing like Yorick \cite{munro1995using} as well as commercial languages and environments like IDL and Matlab.  They also frequently developed code from scratch to solve their own or their colleagues' problems, often in low-level languages that preceeded Python like as Fortran \fixme{Cite https://www.netlib.org/} and C.  It is also hard to imagine that scientific Python was not in part influenced by its contemporaries, such as ROOT.

@@

% https://ras.ac.uk/sites/default/files/2020-01/Group%20Award%20-%20Astropy.pdf
% https://ras.ac.uk/news-and-press/news/leading-astronomers-and-geophysicists-honoured-ras-bicentenary-year-0

This culture of using best practices for producing reliable scientific software has been eagerly adopted by the ecosystem of libraries that build on NumPy.  For example, in a recent award given by the Royal Astronomical Society to AstroPy, they state:
\begin{quotation}
\noindent\emph{The Astropy Project has provided hundreds of junior scientists with experience in professional-standard software development practices including use of version control, unit testing, code review and issue tracking procedures. This is a vital skillset for modern researchers that is often missing from formal university education in physics or astronomy.}
\end{quotation}
Community members explicitly work to address this lack of formal education through formal courses and workshops\cite{wilson-software-carpentry,hannay-scientific-software-survey,millman2018teaching}\fixme{cite https://scipy-school.org/}.
